---
title: "Using Data Science to Make a Christmas Playlist that Doesn't Suck"
author: "Brad Stieber"
date: "November 22, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
```

In this post, I'm going to using data from Spotify to make a Christmas playlist that you'll actually like. We're going to employ a few standard data science techniques to __import__ data, __explore__ it, and finally __model__ the data.

Here are the steps we're going to take

1. Pull data into `R` using the [__`spotifyr`__](https://www.rcharlie.com/spotifyr/) package
1. Clean up the data using functions from the [__`caret`__](http://caret.r-forge.r-project.org/) package
1. Use principal components analysis to reduce the dimensionality of our data
1. Apply k-nearest neighbor and logistic regression modeling techniques to find Christmas songs that don't suck
1. Use our findings to create a playlist that you won't feel like turning off after ten minutes

Here are the packages we'll need:

```{r}
library(spotifyr)
library(tidyverse)
library(scales)
library(caret)
library(ggrepel)
library(RANN)
theme_set(theme_bw())
```


## Getting data from Spotify

The `spotifyr` package makes it really easy to grab data from Spotify. You can find a pretty good tutorial to get up-and-running from the [package's website](https://www.rcharlie.com/spotifyr/).

In the code below, we're going to connect to Spotify, and then pull all of my playlists. I created two playlists for this analysis. The first is called __holiday_playlist__, and it's a collection of 21 songs that I thought were pretty good. Then, we're going to pull in another playlist called __spotify_holiday_playlists__, which is a collection of 300+ holiday songs coming from a variety of Spotify holiday playlists (you can find the list of playlists I used [on my GitHub](https://github.com/bgstieber/files_for_blog/blob/master/make-christmas-playlist/spotify_playlists.txt)).

```{r}
# connect
access_token <- get_spotify_access_token()
# get my playlists
my_playlists <- get_user_playlists('12102534356') 
# get playlist and playlist track info
## my curated list
holiday_playlist <- my_playlists  %>%
  filter(playlist_name == 'holiday_playlist') %>%
  get_playlist_tracks()

holiday_playlist_features <- holiday_playlist %>%
  get_track_audio_features()

## collection of spotify songs
spotify_holiday <- my_playlists %>%
  filter(playlist_name == 'spotify_holiday_playlists') %>%
  get_playlist_tracks() %>%
  # filter duplicates
  filter(!duplicated(.[c('track_name', 'artist_name')] %>%
           mutate_all(toupper)))

spotify_holiday_features <- spotify_holiday %>%
  get_track_audio_features()
```



