---
title: "Mapping COVID Mask Usage in R"
author: "Brad Stieber"
date: "8/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```


The coronavirus pandemic is still not under control [across the United States](https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html). One of the [easiest ways](http://files.fast.ai/papers/masks_lit_review.pdf) to [protect you and those around you](https://www.washingtonpost.com/business/2020/06/17/masks-salons-missouri/) is by [wearing a mask](http://ftp.iza.org/dp13319.pdf) when [you leave the house](https://www.healthaffairs.org/doi/10.1377/hlthaff.2020.00818). The New York Times performed a [survey](https://www.nytimes.com/interactive/2020/07/17/upshot/coronavirus-face-mask-map.html) to investigate mask usage across the US. In this post, I'm going to explore that data a bit, and then end with a cautionary tale of calculating percentages with a small sample.


## The Data

In a separate script, I've collected the data we'll need for this analysis.

```{r echo = FALSE}
model_data <- 
```


## Let's Make a Map

## Let's Make Another Map


## Calculating Percentages with a Small Sample

```{r cache = TRUE}
experiment <- replicate(10000, expr = c("n10" = mean(rbinom(10, 1, .5)), 
                                        "n100" = mean(rbinom(100, 1, .5)), 
                                        "n1000" = mean(rbinom(1000, 1, .5))))

experiment_df <- as_tibble(t(experiment))
```

An interesting quirk of this data is the relationship between the population of a county and the % of respondents frequently or always wearing a mask. 

< plot here >

We see a slightly positive relationship between population and mask usage. In my opinion, the more interesting element of the chart is that the points form somewhat of a fan shape. 

## Wrapping Up


## Links