---
title: "Mapping COVID Mask Usage in R"
author: "Brad Stieber"
date: "8/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, warning = FALSE}
library(tidyverse)
```


The coronavirus pandemic is still not under control [across the United States](https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html). One of the [easiest ways](http://files.fast.ai/papers/masks_lit_review.pdf) to [protect you and those around you](https://www.washingtonpost.com/business/2020/06/17/masks-salons-missouri/) is by [wearing a mask](http://ftp.iza.org/dp13319.pdf) when [you leave the house](https://www.healthaffairs.org/doi/10.1377/hlthaff.2020.00818). The New York Times performed a [survey](https://www.nytimes.com/interactive/2020/07/17/upshot/coronavirus-face-mask-map.html) to investigate mask usage across the US. In this post, I'm going to explore that data a bit, and then end with a cautionary tale of calculating percentages with a small sample.


## The Data

In a separate [script](https://github.com/bgstieber/files_for_blog/blob/master/covid-mask-map/getdata.R), I've collected [the data](https://github.com/bgstieber/files_for_blog/raw/master/covid-mask-map/covid19_model_data.csv) we'll need for this analysis. The data comes from a variety of sources: the [NYtimes survey](https://github.com/nytimes/covid-19-data/raw/master/mask-use/mask-use-by-county.csv), [covid data](https://usafacts.org/visualizations/coronavirus-covid-19-spread-map/), [2016 election results](https://github.com/tonmcg/US_County_Level_Election_Results_08-16/raw/master/2016_US_County_Level_Presidential_Results.csv), and [the census](https://walker-data.com/tidycensus/index.html).

```{r echo = FALSE, message=FALSE, cache=TRUE}
l1 <- "https://github.com/bgstieber/files_for_blog/raw/master/"
l2 <- "covid-mask-map/covid19_model_data.csv"
model_data <- read_csv(paste0(l1, l2))
```

```{r}
head(model_data)
```

Here are the columns in that data:

- __COUNTYFP__, __fips_numeric__, __county_name__: county-level data
- __freq_always__: % "frequently or always wear a mask in public when you expect to be within six feet of another person
- __deaths_per_100K__: as of July 1st, # of covid-19 deaths per 100K people in the county
- __deaths_per_cases__: as of July 1st, # of covid-19 deaths per covid-19 cases in the county
- __trump_ratio_clinton__: the ratio of Trump's percentage vote to Clinton's ($\%Trump/\%Cllinton$)
- __inc__: per capita income in the county
- __edu_pct__: % of county with a bachelor's or higher
- __population_density__: population density of the county
- __females_65_plus_pct__: % of the female population aged 65 or older in the county
- __state_abbr__: state code 



## Let's Make a Map

## Let's Make Another Map


## Calculating Percentages with a Small Sample

```{r cache = TRUE}
experiment <- replicate(10000, expr = c("n10" = mean(rbinom(10, 1, .5)), 
                                        "n100" = mean(rbinom(100, 1, .5)), 
                                        "n1000" = mean(rbinom(1000, 1, .5))))

experiment_df <- as_tibble(t(experiment))
```

An interesting quirk of this data is the relationship between the population of a county and the % of respondents frequently or always wearing a mask. 

< plot here >

We see a slightly positive relationship between population and mask usage. In my opinion, the more interesting element of the chart is that the points form somewhat of a fan shape. 

## Wrapping Up


## Links