---
title: "Using Cosine Similarity to Make Recommendations in R"
author: "Brad Stieber"
date: "December 29, 2018"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

Recommendation engines impact a large amount of our online lives. The content we watch on Netflix, the products we purchase on Amazon, and even [the homes we buy](https://www.redfin.com/blog/2013/09/the-end-of-search.html) are all served up using recommendation engines. In this post, I'll run through one of the key metrics used in recommendation engines: the cosine similarity.

## Introduction

## The Math: Cosine Similarity

## The Code: Calculating Cosine Similarity in `R`

## The Data: Recommending Songs

We use data from the [Million Song Dataset](https://labrosa.ee.columbia.edu/millionsong/). We're working with a subset of the data (only 10K songs), but that should be enough for us.

```{r echo = TRUE}
library(tidyverse)

play_data <- "https://static.turi.com/datasets/millionsong/10000.txt" %>%
  read_tsv(col_names = c('user', 'song_id', 'plays'))

song_data <- 'https://static.turi.com/datasets/millionsong/song_data.csv' %>%
  read_csv()

all_data <- play_data %>%
  group_by(user, song_id) %>%
  summarise(plays = sum(plays, na.rm = TRUE)) %>%
  inner_join(song_data)
```

Here are the first few rows of the data:

```{r}
head(all_data)
```

## Conclusion

Links:

- [Wikipedia: Cosine Similarity](https://en.wikipedia.org/wiki/Cosine_similarity)
- [Machine Learning :: Cosine Similarity for Vector Space Models (Part III)](http://blog.christianperone.com/2013/09/machine-learning-cosine-similarity-for-vector-space-models-part-iii/)
- [Introduction to Collaborative Filtering](https://hackernoon.com/introduction-to-recommender-system-part-1-collaborative-filtering-singular-value-decomposition-44c9659c5e75)